\chapter*{Acknowledgments}
I would like to thank many people who helped me carry out my research and 
introduced me to \emph{Machine Learning}.

First, I must thank my primary advisor, Reguly István, who pushed me through the stages of understanding the main principles. 
From week to week, we analyzed the results and discussed about further experiments.
He taught me how should I plan a larger project and motivated to implement my own framework.
He read the articles which I was struggling with, and provided literature which I could not find on my own.
In the developing phase of the library, he pointed out and explained many flaws of the current implementation, and suggested popular design patterns, fitting my concepts.

I must also thank my external supervisor, Józsa Csaba, who helped me to summarize my current achievements and encouraged me to dive into larger and larger projects.
He recognized which of my results were interesting and worth further investigation and in which direction I should continue my further studies.
He introduced me to applied techniques of state-of-the-art Neural Networks, and how they can be simplified and applied.
When I came up with ambitious long-term plans, he helped me to understand which ideas would work and how should I start actually implementing them.

I would like to thank Emese Zomborszki, and many others who listened to my explanations.
Without them I could not write down the description of my recent studies in such a way,
that those who are not involved in the area of Machine Learning can also recognize and understand the importance of it.

Finally I have to thank the authors of the cited literature,
especially the writers of on-line tutorials Michael Nielsen, Andrej Karpathy and Andrew Gibiansky.
Their results inspired me to join the community of researchers and motivate us to contribute to the science of Artificial Intelligence.